---
title: "NOAAStormAnalysis"
author: "SungChan Lee"
date: "Feb 8 2016"
output: html_document
---

## Type of events most harful with respect to polulation health

### Synopsis
*NOAA(National Oceanic and Atmospheric Advministration) storm database has characteristics of major storms and weather events in the United States.   The goal of this analysis is to find out what is the most health threatining event and what event has most serious economic effect.   To find out the event affect on the population heath, fatality and injury data is summed by event type.   To figure out the economic effect, numeric calcuation has been performed to convert exponential code into actual values.   Three plots are provided to visualize the analysis result.   The result shows Tonado is the most significant event for fatalities and injuries.   On contrary the highst event for economic damage is Flood (Tonado is the 3rd event)*

### Data Processing

This line is requred to prevent following error   
"Warning message:In scan(file, what, nmax, sep, dec, quote, skip, nlines, na.strings,  : 
EOF within quoted string""

```{r, echo=TRUE}

library(ggplot2)
Sys.setlocale("LC_ALL", "English")
```

When we download or unzip data, it is required to check if there is already data to avoid repeating unnecessary work.

```{r, echo=TRUE}
#cache zipped file
if(!file.exists("stormdata.csv.bz2")) {
  download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2", destfile = "stormdata.csv.bz2")
}

#cache uncompressed file
if(!file.exists("stormdata.csv")) 
  bunzip2("stormdata.csv.bz2", remove="FALSE")

#cache data
if(!exists("raw_storm_data")) 
  raw_storm_data <- read.csv("stormdata.csv")
```

### Filter out unused columns through this project
```{r, echo=TRUE}

#filter necessary columns
storm_data <- raw_storm_data[,c('EVTYPE','FATALITIES','INJURIES', 'PROPDMG', 'PROPDMGEXP', 'CROPDMG', 'CROPDMGEXP')]

#print (head(storm_data))
```


### Count fatalities and injuries for each event type
```{r, echo=TRUE}
#count fatalities per event type
fatal <- aggregate(FATALITIES~EVTYPE, data=storm_data, FUN=sum, na.rm=TRUE, na.action=NULL)
fatal <- fatal[order(fatal$FATALITIES, decreasing=TRUE), ]
fatal.top5 <- head(fatal, 5)
fatal.top5
barplot(fatal.top5$FATALITIES, names.arg = fatal.top5$EVTYPE, xlab="Event Type", ylab="Fatalities", cex.names=0.7, main="Top5 event with high fatalities")

#count injury per event type
injury <- aggregate(INJURIES~EVTYPE, data = storm_data, FUN=sum, na.rm=TRUE, na.action=NULL)
injury <- injury[order(injury$INJURIES, decreasing=TRUE), ]
injury.top5 <- head(injury, 5)
injury.top5
barplot(injury.top5$INJURIES, names.arg = injury.top5$EVTYPE, xlab="Event Type", ylab="Injuries", cex.names=0.7, main="Top5 event with high Injuries")

```

###Result
<span style="color:red">**Barplot shows that Tonado has the most significant effect on the population health** </span>

## Type of event have the greatest economic consequences

First of all, we should check how damage data is described
```{r, echo=TRUE}
## check what levels are 
print(unique(storm_data$PROPDMGEXP))
print(unique(storm_data$CROPDMGEXP))
```

According to the observation, we should convert exponential code into exponential value   
```{r, echo=FALSE}
storm_data$PROPDMGEXP <- sub(pattern="^$|[-?+]", replacement="0", x=storm_data$PROPDMGEXP)
storm_data$PROPDMGEXP <- sub(pattern="[H]", replacement="2", ignore.case= TRUE,  x=storm_data$PROPDMGEXP)
storm_data$PROPDMGEXP <- sub(pattern="[K]", replacement="3", ignore.case= TRUE,  x=storm_data$PROPDMGEXP)
storm_data$PROPDMGEXP <- sub(pattern="[M]", replacement="6", ignore.case= TRUE,  x=storm_data$PROPDMGEXP)
storm_data$PROPDMGEXP <- sub(pattern="[B]", replacement="9", ignore.case= TRUE,  x=storm_data$PROPDMGEXP)

storm_data$CROPDMGEXP <- sub(pattern="^$|[?]", replacement="0", x=storm_data$CROPDMGEXP)
storm_data$CROPDMGEXP <- sub(pattern="[K]", replacement="3", ignore.case= TRUE,  x=storm_data$CROPDMGEXP)
storm_data$CROPDMGEXP <- sub(pattern="[M]", replacement="6", ignore.case= TRUE,  x=storm_data$CROPDMGEXP)
storm_data$CROPDMGEXP <- sub(pattern="[B]", replacement="9", ignore.case= TRUE,  x=storm_data$CROPDMGEXP)
```

Let's see exponential code was replaced correctly   
```{r, echo=TRUE}
print(unique(storm_data$PROPDMGEXP))
print(unique(storm_data$CROPDMGEXP))
```

### We should calculate total damage 
```{r, echo=TRUE}
#calculate damage
storm_data$TOTALDMG <- as.numeric(storm_data$PROPDMG) * (10 ^ as.numeric(storm_data$PROPDMGEXP)) + as.numeric(storm_data$CROPDMG) * (10 ^ as.numeric(storm_data$CROPDMGEXP))
```

### Aggregate damage data by summation according to event type
```{r, echo=FALSE}
#aggregate
storm_data.damage <- aggregate(TOTALDMG~EVTYPE, data = storm_data, FUN=sum, na.rm=TRUE, na.action=NULL)
#print(head(storm_data.damage, 20))

#sort
storm_data.damage <- storm_data.damage[order(storm_data.damage$TOTALDMG, decreasing=TRUE), ]
```

### Bar plot top5 
```{r, echo=FALSE}

storm_data.damage.top5 = head(storm_data.damage, 5)
#storm_data.damage.top5
barplot(storm_data.damage.top5$TOTALDMG, names.arg = storm_data.damage.top5$EVTYPE, xlab="Event Type", ylab="Total Damage", cex.names=0.5, main="Top5 event with high damage")
```

### Result
<span style="color:red">**Flood is the event with the highst damage**</span>
